---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# azores.bkw

<!-- badges: start -->
<!-- badges: end -->

The goal of `{azores.bkw}` is to provide an R Compendium Accompanying ...

## Installation

``` r
# install.packages("pak")
pak::pak("patterninstitute/azores.bkw")
```
## Sightings

A data set of sightings of beaked whales in the Azores, during the period of 2012 and 2018:

```{r sightings}
library(azores.bkw)
library(ggplot2)
library(dplyr)
library(CAOP.RAA.2024)
library(azores.rorquals)

bkw.sightings
```

## Beaked whales occurrences


The `bkw_occurrence` dataset is derived from `presences_sf2` from package `azores.rorquals` through
the following steps:

1. Only observations from the `rps` dataset were kept

1. Observations with missing (`NA`) values for any climate or topographical
variable are removed.

2. Observations of the target species---Sowerby’s beaked whale (*Mesoplodon bidens*), 
True’s beaked whale *Mesoplodon mirus*, Blanville’s beaked whale (*Mesoplodon densirostris*), 
Gervais’s beaked whale (*Mesoplodon europaeus*), Northern bottlenose whales (*Hyperoodon ampullatus*),
and Cuvier’s beaked whale (*Ziphius cavirostris*)---are classified as "presence" in the
class column, while all other species are labeled as "absence".

3. Observations that could simultaneously represent both presences of target
species and absences of others are excluded.

4. Observations located on land are removed. This issue may arise due to rasterization
effects in climate and topographical data, when merging locations that are very close to 
the shore (< 500 m).

```{r fig.width=10, fig.height=8}
bkw_occurrence

colours <- c(
  Balaenopteridae = "#0F00FF",
  Physeteridae = "#F404E2",
  Ziphiidae = "#FF7A00",
  Delphinidae  = "#E8E8E8AF",
  Kogiidae = "#00E000",
  Balaenidae = "#FFEE00AF",
  Phocoenidae = "#04B3F4"
)

islands_triangle <- CAOP.RAA.2024::districts() |>
  dplyr::filter(district %in% c("Ilha de São Jorge",
                         "Ilha do Pico",
                         "Ilha do Faial"))

bkw_occurrence |>
  dplyr::mutate(season = azores.rorquals::as_season(date)) |>
  dplyr::left_join(cetaceans(), by = "species") |>
  tidyr::drop_na(family) |>
  ggplot() +
  geom_sf(data = islands_triangle, fill = "black") +
  geom_sf(mapping = aes(col = family), size = 0.1) +
  scale_color_manual(values = colours, name = "Cetacean Family") +
  guides(colour = guide_legend(override.aes = list(size = 3))) +
  facet_wrap(vars(season)) +
  theme_minimal() +
  theme(legend.title = element_text(hjust = 0.5),
        strip.text = element_text(size = 14))
```
